---
title: "Genome-wide allele-specific splicing"
author: "XSun"
date: "2026-09-28"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      fig.width=7, fig.height=5)
library(data.table)
library(ggplot2)
library(ggrepel)

gene_tp <- fread("docs/assets/roc_gene_tp_signals.tsv")
gene_wide <- fread("docs/assets/roc_gene_signals_wide.tsv")
fmt <- function(n) format(n, big.mark=",", trim=TRUE)
```

We ran [isoLASER](https://github.com/gxiaolab/isoLASER) jointly across four
donors on all 22 autosomes, at each of the three timepoints (0, 1, and 6 hours
after stimulation). isoLASER phases reads onto haplotypes using heterozygous
variants it discovers in the RNA, then tests whether the two copies of each gene
splice differently — allele-specific splicing, reported as **dPSI** (the
difference in Percent Spliced In between haplotypes).

The model used is `--platform Nanopore`, which loads the default
`GM12878_ONT_ROC` classifier. A second ONT classifier (`GTEx_ONT_F1`) ships
with isoLASER but is currently commented out upstream; we tested both and found
F1 recovers more heterozygous sites, but use the default here pending
clarification from the developers.

# Filtering

Four filters were applied to the raw isoLASER cis calls:

1. **Effect size:** |dPSI| >= 0.10. Without this, many statistically significant
   calls have dPSI of 0.01–0.02, driven by high read depth rather than a
   meaningful splicing difference.

2. **Clustered variants removed.** Real heterozygous SNPs occur roughly once per
   kilobase. When multiple variants cluster within 30 bp, they are likely
   sequencing errors in difficult sequence. 34% of all variants fall in such
   clusters.

3. **Per-donor allelic balance.** At least one donor must have allelic balance
   0.25–0.75 and read depth >= 10 at the phasing variant. This is checked
   per donor, not averaged across donors — an average can hide a case where
   no individual donor is heterozygous (e.g., one reference-only and one
   alternate-only donor averaging to 0.5).

4. **Minimum exonic part length >= 10 bp.** The exon model splits overlapping
   annotations into non-overlapping parts, which can create fragments of 1–3 bp.
   Allelic balance measured over such short regions is unreliable.

# Signal counts

After filtering: **`r fmt(uniqueN(gene_wide$gene))` genes** with at least one
allele-specific splicing event. Each gene is represented by its strongest
(gene, variant, exonic part) combination, tracked as a fixed unit across all
three timepoints.

```{r genes_per_tp, fig.height=4}
tp_summary <- gene_tp[has_sig==TRUE, .(n_genes=uniqueN(gene)), by=tp]
tp_summary[, tp := factor(tp, levels=c("0hr","1hr","6hr"))]

ggplot(tp_summary, aes(tp, n_genes)) +
  geom_col(fill=c("#888888","#E69F00","#D55E00"), width=0.6) +
  geom_text(aes(label=n_genes), vjust=-0.3, size=4) +
  labs(x="Timepoint", y="Genes with cis signal") +
  ylim(0, max(tp_summary$n_genes)*1.15) +
  theme_minimal(base_size=14)
```

# Timepoint patterns

For each gene, we classified its signal across the three timepoints.

```{r tp_patterns, fig.height=4}
pat <- gene_wide[, .N, by=pattern]
pat[, pattern := factor(pattern, levels=c("stimulation_dependent",
                                           "post_stim_only",
                                           "all_timepoints",
                                           "lost_after_stim","other"))]
setorder(pat, pattern)

ggplot(pat, aes(pattern, N)) +
  geom_col(fill="#4477AA", width=0.6) +
  geom_text(aes(label=N), vjust=-0.3, size=3.5) +
  labs(x="", y="Number of genes") +
  theme_minimal(base_size=13) +
  theme(axis.text.x=element_text(angle=30, hjust=1))
```

```{r tp_table}
pat_desc <- data.frame(
  Pattern=c("stimulation_dependent", "post_stim_only",
            "all_timepoints","lost_after_stim"),
  Description=c(
    "Tested at 0 hr with no signal; signal appears after stimulation",
    "Not tested at 0 hr (no reads or no variant); signal only at 1 or 6 hr — absence at baseline is not established",
    "Signal at all three timepoints (constitutive)",
    "Signal at 0 hr but absent at one or both stimulated timepoints"),
  Genes=c(sum(gene_wide$pattern=="stimulation_dependent"),
          sum(gene_wide$pattern=="post_stim_only"),
          sum(gene_wide$pattern=="all_timepoints"),
          sum(gene_wide$pattern=="lost_after_stim")))
knitr::kable(pat_desc)
```

Of the genes with a post-stimulation signal, **`r sum(gene_wide$pattern=="stimulation_dependent")`
were tested at baseline with no signal** — these are the strongest candidates
for stimulation-dependent allele-specific splicing. The remaining
`r sum(gene_wide$pattern=="post_stim_only")` were not tested at 0 hr at all
(the gene was not expressed, had no heterozygous variant, or failed a filter),
so baseline absence is not established.

# Top signals

```{r top30, fig.height=9, fig.width=7}
top_genes <- gene_wide[order(-max_signal)][1:30, gene]

top_data <- gene_tp[gene %in% top_genes & has_sig == TRUE]
top_data[, tp := factor(tp, levels=c("0hr","1hr","6hr"))]
top_data[, gene := factor(gene, levels=rev(top_genes))]

ggplot(top_data, aes(tp, gene, size=abs_dpsi, color=abs_dpsi)) +
  geom_point() +
  scale_size_continuous(range=c(1,8), name="|dPSI|") +
  scale_color_gradient(low="#DDDDDD", high="#CC0000", name="|dPSI|") +
  labs(x="Timepoint", y="") +
  theme_minimal(base_size=12) +
  theme(panel.grid.major=element_line(color="#EEEEEE"))
```

Top 30 genes ranked by maximum |dPSI|. Each gene is represented by a single
(variant, exonic part) combination tracked across all three timepoints. Missing
dots mean that combination had no signal at that timepoint.

```{r top_table}
top_tab <- gene_wide[order(-max_signal)][1:30]
knitr::kable(top_tab[, .(gene, hr0, hr1, hr6, pattern)],
             digits=3, col.names=c("Gene","0 hr","1 hr","6 hr","Pattern"))
```

# Neuronal and activity-regulated genes

```{r named_genes, fig.height=6, fig.width=8}
ieg <- c("FOS","FOSB","FOSL1","FOSL2","JUN","JUNB","JUND",
         "EGR1","EGR2","EGR3","EGR4","NR4A1","NR4A2","NR4A3",
         "ARC","BDNF","NPAS4","SIK1","ATF3","GADD45B","GADD45G",
         "HOMER1","IER2","IER5","BTG2","DUSP1","DUSP5","DUSP6",
         "NFKBIA","NFKBIZ","CREM","CYR61","CTGF","TIPARP","PPP1R15A",
         "ZFP36","ZFP36L1","BHLHE40","BHLHE41","KLF2","KLF4",
         "RGS2","SERTAD1","CITED2")
neuronal <- c("NRXN1","NRXN2","NRXN3","NLGN1","NLGN2","NLGN3",
              "GRIN1","GRIN2A","GRIN2B","GRIA1","GRIA2","GRIA3","GRIA4",
              "DLG4","SHANK1","SHANK2","SHANK3","CAMK2A","CAMK2B",
              "SYN1","SYT1","SNAP25","CACNA1A","CACNA1B","CACNA1C","CACNA1E",
              "SCN1A","SCN2A","SCN8A","CELF4","RBFOX1","RBFOX2","RBFOX3",
              "NOVA1","NOVA2","PTBP2","ANK3","MAPT","APP","DSCAM",
              "CNTNAP2","CADM1","TTLL7","CLDND1","MEF2C")

gene_all <- gene_tp[has_sig==TRUE, .(max_dpsi=max(abs_dpsi), mean_cov=median(COV, na.rm=TRUE)), by=gene]
gene_all <- merge(gene_all, gene_wide[, .(gene, pattern)], by="gene")
gene_all[, label_cat := fcase(
  gene %in% ieg, "Activity-regulated",
  gene %in% neuronal, "Neuronal",
  default = "Other"
)]
gene_all[, show_label := label_cat != "Other" & max_dpsi >= 0.10]

ggplot(gene_all, aes(mean_cov, max_dpsi, color=label_cat)) +
  geom_point(data=gene_all[label_cat=="Other"], alpha=0.3, size=1.5) +
  geom_point(data=gene_all[label_cat!="Other"], alpha=0.8, size=3) +
  geom_text_repel(data=gene_all[show_label==TRUE],
                  aes(label=gene), size=3.5, max.overlaps=20, show.legend=FALSE) +
  scale_color_manual(values=c("Activity-regulated"="#D55E00",
                               "Neuronal"="#0072B2", "Other"="#BBBBBB"), name="") +
  scale_x_log10() +
  labs(x="Median coverage (reads)", y="Max |dPSI| across timepoints") +
  theme_minimal(base_size=13)
```

Labeled genes are from curated lists of activity-regulated immediate-early genes
(orange) and neuronal-function genes (blue).

```{r named_table}
named <- gene_wide[gene %in% c(ieg, neuronal)]
named[, category := fifelse(gene %in% ieg, "Activity-regulated", "Neuronal")]
named <- named[max_signal > 0]
setorder(named, -max_signal)
knitr::kable(named[, .(gene, category, hr0, hr1, hr6, pattern)],
             digits=3,
             col.names=c("Gene","Category","0 hr","1 hr","6 hr","Pattern"))
```

Most classic immediate-early genes (FOS, JUN, EGR1, ARC) are not testable by
isoLASER — they are either single-exon genes, have no heterozygous variant in
the four donors, or have too few spliced reads.

# Top neuronal gene plots

```{r gene_plots, fig.height=4, fig.width=7, results='asis'}
# Show each gene's selected (variant, exonic_part) across timepoints,
# plus any other events faceted separately so they are not mixed.
event_data <- fread("/project/xinhe/xsun/long_read_RNA/9.isolaser_results/out/roc_event_wide.tsv.gz")
focus_genes <- c("GRIA2","ANK3","CACNA1E","SCN2A","CAMK2B","SYT1","NFKBIA")
focus_genes <- focus_genes[focus_genes %in% event_data$gene]

for(g in focus_genes) {
  gd <- event_data[gene == g & pattern != "other"]
  if(nrow(gd) == 0) next
  # reshape to long
  gd_long <- melt(gd, id.vars=c("gene","variant_qual","exonic_part","pattern"),
                  measure.vars=c("dpsi_0hr","dpsi_1hr","dpsi_6hr"),
                  variable.name="tp", value.name="dpsi")
  gd_long[, tp := factor(sub("dpsi_","",tp), levels=c("0hr","1hr","6hr"))]
  gd_long <- gd_long[!is.na(dpsi)]

  # label each event
  gd_long[, event_id := paste0(
    sub(".*:(\\d+):.*","\\1", variant_qual), " / ",
    sub("chr[0-9]+:(\\d+)-(\\d+):.*","\\1-\\2", exonic_part))]

  # pick up to 6 strongest events
  top_events <- gd[order(-max_abs_dpsi)][1:min(.N, 6), .(variant_qual, exonic_part)]
  gd_long <- gd_long[top_events, on=.(variant_qual, exonic_part)]

  p <- ggplot(gd_long, aes(tp, dpsi)) +
    geom_hline(yintercept=0, lty=2, color="grey60") +
    geom_point(size=3, color="#0072B2") +
    geom_line(aes(group=1), color="#0072B2", alpha=0.5) +
    facet_wrap(~event_id, scales="free_y") +
    labs(x="Timepoint", y="dPSI", title=g) +
    theme_minimal(base_size=12)
  print(p)
  cat("\n\n")
}
```

Each panel shows one (variant, exonic part) combination tracked across
timepoints. The variant position and exonic part coordinates are in the facet
label (variant position / part start-end). This ensures each panel represents a
single splicing event, not a mixture.

# Interpretation

This analysis identifies **`r nrow(gene_wide)` genes** with allele-specific
splicing after quality filtering. Each gene is represented by a fixed
(variant, exonic part) combination tracked across all three timepoints, so
the timepoint comparison is always between the same splicing event.
**`r sum(gene_wide$pattern=="stimulation_dependent")` genes** were tested at
baseline with no signal and then showed a signal after stimulation — these are
the strongest candidates for stimulation-dependent allele-specific splicing.
Another `r sum(gene_wide$pattern=="post_stim_only")` genes have post-stimulation
signals but were not tested at baseline, so stimulation dependence is not
established for those.

The strongest neuronal hits are GRIA2 (AMPA receptor subunit, 6 hr only),
ANK3 (spectrin adapter, all timepoints), CACNA1E (calcium channel, 6 hr only),
and SCN2A (sodium channel, 6 hr only). These are genes with known alternative
splicing in neurons, making the allele-specific signals plausible.

Caveats:

- **No sensitivity estimate.** This analysis starts from isoLASER's calls and
  does not count variants or genes that isoLASER missed. The default Nanopore
  model is known to miss ~36% of heterozygous sites (see the
  [variant validation](isolaser_variants.html) page), which reduces the number
  of testable genes.
- **Four donors.** With n = 4, a signal requires a heterozygous variant in at
  least one donor. Rare variants are not testable.
- **No multiple-testing correction** has been applied. The p-values from
  isoLASER's joint model are shown as-is.

Related pages: [variant validation](isolaser_variants.html),
[junction screen](splicing_candidates.html),
[isoLASER setup](isolaser.html).
