---
title: "Genome-wide allele-specific splicing"
author: "XSun"
date: "2026-09-28"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      fig.width=7, fig.height=4)
library(data.table)
library(ggplot2)
library(DT)
gene_tp <- fread("docs/assets/roc_gene_tp_signals.tsv")
gene_wide <- fread("docs/assets/roc_gene_signals_wide.tsv")
events <- fread("docs/assets/roc_event_wide.tsv.gz")
passing <- fread("docs/assets/roc_cis_events_filtered.tsv.gz")
sig <- function(x) !is.na(x) & x == TRUE
fmt <- function(n) format(n, big.mark=",", trim=TRUE)
tp_levels <- c("0hr", "1hr", "6hr")
pattern_names <- c(detected_after_stim="Detected after stimulation",
  post_stim_only="Baseline unavailable", all_timepoints="All three timepoints",
  detected_at_baseline_only="Baseline only", other="Other detection patterns")
pass_times <- function(d) vapply(seq_len(nrow(d)), function(i) {
  keep <- sig(unlist(d[i, .(passes_0hr, passes_1hr, passes_6hr)]))
  paste(c("0", "1", "6")[keep], collapse=", ")
}, character(1))
effect_table <- function(d) {
  data.table(Gene=d$gene,
    `0 hr`=round(100*abs(d$dpsi_0hr),1),
    `1 hr`=round(100*abs(d$dpsi_1hr),1),
    `6 hr`=round(100*abs(d$dpsi_6hr),1),
    `Passing timepoints (hr)`=pass_times(d),
    Pattern=unname(pattern_names[d$pattern]))
}
interactive_table <- function(d) DT::datatable(as.data.frame(d), rownames=FALSE,
  filter="top", extensions="Buttons", options=list(
    pageLength=10, lengthMenu=c(10,25,50,100), scrollX=TRUE,
    order=list(), dom="Blfrtip", buttons=c("copy","csv")),
  caption="Search, filter, or click a column heading to sort; export the filtered table as CSV.")
neuronal_genes <- c("NRXN1","NRXN2","NRXN3","NLGN1","NLGN2","NLGN3",
  "GRIN1","GRIN2A","GRIN2B","GRIA1","GRIA2","GRIA3","GRIA4",
  "DLG4","SHANK1","SHANK2","SHANK3","CAMK2A","CAMK2B","SYN1","SYT1","SNAP25",
  "CACNA1A","CACNA1B","CACNA1C","CACNA1E","SCN1A","SCN2A","SCN8A",
  "CELF4","RBFOX1","RBFOX2","RBFOX3","NOVA1","NOVA2","PTBP2","ANK3","MAPT",
  "APP","DSCAM","CNTNAP2","CADM1","TTLL7","MEF2C","DCLK1","DTNBP1","CADPS",
  "DOCK7","DMXL2","SIGMAR1","TMEM106B","FNDC5","ELAVL4","PARK7")
activity_genes <- c("FOS","FOSB","FOSL1","FOSL2","JUN","JUNB","JUND",
  "EGR1","EGR2","EGR3","EGR4","NR4A1","NR4A2","NR4A3","ARC","BDNF","NPAS4",
  "SIK1","ATF3","GADD45B","GADD45G","HOMER1","IER2","IER5","BTG2","DUSP1",
  "DUSP5","DUSP6","NFKBIA","NFKBIZ","CREM","TIPARP","PPP1R15A","ZFP36",
  "BHLHE40","BHLHE41","KLF2","KLF4","RGS2","SERTAD1","CITED2")
neuronal_events <- copy(events[gene %in% c(neuronal_genes,activity_genes)])
priority_patterns <- c("detected_after_stim","all_timepoints",
  "detected_at_baseline_only","other","post_stim_only")
neuronal_events[, priority := match(pattern,priority_patterns)]
setorder(neuronal_events,priority,-max_passing_dpsi)
neuronal_selected <- neuronal_events[, .SD[1], by=gene]
setorder(neuronal_selected,priority,-max_passing_dpsi)
candidate_table <- function(d) {
  ans <- effect_table(d)
  ans[, `Usable baseline` := ifelse(sig(d$usable_0hr),"Yes","No")]
  ans[, Variant := d$variant_qual]
  ans[, `Exonic part` := d$exonic_part]
  ans
}
```

We screened **22 autosomes, four donors, and three timepoints (0, 1, and
6 hours)** using isoLASER's default Nanopore model (`GM12878_ONT_ROC`).
The filtered joint analysis identifies **`r fmt(nrow(gene_wide))` genes** with
candidate allele-associated splicing signals. These are candidates for
follow-up; within-donor allele-specific splicing and stimulation dependence
have not yet been confirmed.

# How the screen works

At each timepoint, isoLASER jointly evaluates associations between a variant
and an **exonic part**, a segment of the gene's exon structure. We retain its
`cis` calls using these filters:

| Filter | Requirement | Purpose |
|:--|:--|:--|
| Allelic difference | \(\lvert\Delta\mathrm{PSI}\rvert \geq 0.10\) | At least a 10 percentage-point difference |
| Donor eligibility | At least one donor with alternate-allele fraction 0.25–0.75 and depth ≥10 | Require reads from both alleles in an individual donor |
| Part length | ≥10 bases | Exclude very short fragments |
| Variant spacing | No other tested variant within 30 bases | Conservative exclusion of clustered calls; real variants can also be removed |

For timepoint comparisons, we keep the **same gene, variant, and exonic part**.
The representative event for each gene is the one with the largest effect
that passes these filters. A usable timepoint has an output row and at least
one donor meeting the allele-balance and depth criteria; the eligible donor
need not be the same at different timepoints.

## What the numbers mean

For a given part, an allele-specific inclusion rate is:

\[
\mathrm{PSI}_{\mathrm{REF}}=
\frac{\text{reference-allele reads including the part}}
{\text{reference-allele reads including or excluding the part}},
\qquad
\Delta\mathrm{PSI}=\mathrm{PSI}_{\mathrm{REF}}-\mathrm{PSI}_{\mathrm{ALT}}.
\]

For example, 16/20 reference reads versus 6/20 alternate reads gives
80% versus 30%, or an inclusion difference of **50 percentage points**.
This is an illustration, not a measured result from this dataset.

**Sign convention in these outputs:** the installed joint implementation
computes its signed `delta_psi(ref-alt)` from the first count category, which
is the part's **exclusion** category. Thus the reported joint difference is
reference exclusion rate minus alternate exclusion rate; the inclusion
difference has the opposite sign. The absolute effect is unchanged. The
joint calculation uses coverage-normalized counts across donors, rather
than a simple sum of raw reads.

# How many signals were found?

```{r count_table}
knitr::kable(data.frame(
  Unit=c("Passing rows", "Distinct gene–variant–part combinations", "Genes"),
  Count=c(fmt(nrow(passing)), fmt(nrow(events)), fmt(nrow(gene_wide))),
  Meaning=c("One gene–variant–exonic-part association at one timepoint",
    "Associations combined across timepoints", "At least one passing association")))
```

These units are not independent biological events: multiple variants or
adjacent parts may describe the same underlying splicing change.
For example, one gene–variant–part association passing at 0, 1, and 6 hours
contributes **three passing rows**, but only **one distinct association**.

```{r genes_per_tp}
tp_summary <- passing[, .(Genes=uniqueN(gene)), by=tp]
tp_summary[, tp := factor(tp, levels=tp_levels)]
ggplot(tp_summary, aes(tp, Genes, fill=tp)) +
  geom_col(width=0.6, show.legend=FALSE) +
  geom_text(aes(label=Genes), vjust=-0.4, size=4) +
  scale_fill_manual(values=c("0hr"="#888888", "1hr"="#E69F00", "6hr"="#D55E00")) +
  scale_x_discrete(labels=c("0 hr", "1 hr", "6 hr")) +
  scale_y_continuous(expand=expansion(mult=c(0,0.12))) +
  labs(x=NULL, y="Genes with any passing association") +
  theme_minimal(base_size=13)
```

## Detection across timepoints

The following counts describe each gene's **selected representative event**.
They describe filter outcomes, not a statistical test of whether stimulation
changes the allelic effect.

**Usable** means that this exact gene–variant–part combination has a joint
output row at that timepoint, and at least one donor has alternate-allele
fraction **0.25–0.75** with **depth ≥10** for the combination. Usable data
can still fail the `cis` or effect-size criteria. **No usable baseline** can
mean a missing output row, insufficient depth, or no donor meeting the
allele-balance criterion; it does not establish that the gene is unexpressed.

```{r detection_barplot, fig.height=4.5}
detection_counts <- data.table(pattern=names(pattern_names))
detection_counts[, Genes := vapply(pattern,function(p)sum(gene_wide$pattern==p),integer(1))]
detection_counts[, Pattern := factor(unname(pattern_names[pattern]),
  levels=rev(unname(pattern_names)))]
ggplot(detection_counts,aes(Pattern,Genes,fill=pattern)) +
  geom_col(width=0.65,show.legend=FALSE) +
  geom_text(aes(label=Genes),hjust=-0.2,size=4) + coord_flip() +
  scale_fill_manual(values=c(detected_after_stim="#0072B2",post_stim_only="#BBBBBB",
    all_timepoints="#009E73",detected_at_baseline_only="#E69F00",other="#999999")) +
  scale_y_continuous(expand=expansion(mult=c(0,0.12))) +
  labs(x=NULL,y="Genes (selected representative event)") + theme_minimal(base_size=12)
```

```{r detection_table}
patterns <- names(pattern_names)
descriptions <- c(
  "Usable baseline does not pass; passes at 1 or 6 hr",
  "No usable baseline; passes at 1 or 6 hr",
  "Passes at 0, 1, and 6 hr",
  "Passes at 0 hr; usable at 1 and 6 hr but passes at neither",
  "Mixed detection or incomplete usable timepoints")
knitr::kable(data.frame(Pattern=unname(pattern_names),
  Genes=vapply(patterns,function(p)sum(gene_wide$pattern==p),integer(1)),
  Definition=descriptions), row.names=FALSE)
```

**`r sum(gene_wide$pattern=="detected_after_stim")` genes** pass after
stimulation despite a usable, nonpassing baseline. They are a useful starting
set for follow-up. For **`r sum(gene_wide$pattern=="post_stim_only")` genes**,
baseline absence cannot be assessed.

# Neuronal and activity-associated candidates

The table below expands the examples to **`r nrow(neuronal_selected)` genes**
from a manually selected list of neuronal and activity-associated genes.
**`r sum(neuronal_selected$pattern=="detected_after_stim")` have an association
detected after stimulation with a usable, nonpassing baseline.** These appear
first, followed by other baseline-usable patterns; candidates without a usable
baseline appear last. Biological relevance helps prioritize examples but does
not establish a stimulation-dependent allelic effect.

For this section, we search **all passing associations**, rather than only the
strongest event per gene. We select one association per gene by detection
category, then by the largest passing effect within that category. This can
select a different event from the genome-wide representative above. Each
selected variant–part combination remains fixed across timepoints; the
genome-wide counts and detection categories above are unchanged.

The expanded candidates include genes involved in neuronal structure and
secretion: [DCLK1 and dendritic cargo transport](https://pubmed.ncbi.nlm.nih.gov/26758546/),
[CADPS and calcium-regulated vesicle secretion](https://www.ncbi.nlm.nih.gov/gene/8618/),
and [DTNBP1 and synaptic function](https://pubmed.ncbi.nlm.nih.gov/27330059/).

```{r examples_table}
neuronal_table <- candidate_table(neuronal_selected)
neuronal_table[, Category := ifelse(neuronal_selected$gene %in% activity_genes,
  "Activity-associated","Neuronal")]
interactive_table(neuronal_table)
```

Table values are **absolute allelic differences in percentage points**, not
the percentage of transcripts including an exon. `NA` means no joint estimate
was available. Nonpassing measured effects remain visible; the passing
timepoints are listed separately.

## Baseline-usable examples to follow up

The plots show the six strongest passing effects among the selected
**detected-after-stimulation** candidates. No estimate is substituted for
missing data. Read-level inspection and within-donor comparisons are the next
steps for these examples.

```{r example_trends, fig.height=6, fig.width=9}
focus <- neuronal_selected[pattern=="detected_after_stim"][order(-max_passing_dpsi)]
focus <- focus[seq_len(min(6,nrow(focus)))]
plot_data <- melt(focus,id.vars=c("gene","variant_qual","exonic_part"),
  measure.vars=list(dpsi=paste0("dpsi_",tp_levels),
    passes=paste0("passes_",tp_levels),usable=paste0("usable_",tp_levels)),
  variable.name="time_index")
plot_data[, tp := tp_levels[as.integer(time_index)]]
plot_data[, tp := factor(tp,levels=tp_levels)]
plot_data[, gene := factor(gene,levels=focus$gene)]
plot_data[, Status := fifelse(sig(passes), "Passes filters",
  fifelse(sig(usable), "Usable; does not pass", "Does not meet donor eligibility"))]
plot_data <- plot_data[!is.na(dpsi)]
ggplot(plot_data,aes(tp,100*dpsi)) +
  geom_hline(yintercept=0,linetype=2,color="grey65") +
  geom_point(aes(shape=Status,color=Status),size=3) +
  facet_wrap(~gene,ncol=3) +
  scale_shape_manual(values=c("Passes filters"=16,"Usable; does not pass"=1,
    "Does not meet donor eligibility"=4)) +
  scale_color_manual(values=c("Passes filters"="#0072B2",
    "Usable; does not pass"="#777777","Does not meet donor eligibility"="#BBBBBB")) +
  scale_x_discrete(labels=c("0 hr","1 hr","6 hr"),drop=FALSE) +
  labs(x=NULL,y="Joint exclusion difference (REF − ALT, percentage points)",
    shape=NULL,color=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
```

Each panel follows the same association across timepoints on a common vertical
scale; a positive value means more exclusion on the reference allele.
Missing points are missing estimates, and the plot does not imply
that donor-level allele-specific rates have been measured.

```{r example_coordinates}
knitr::kable(focus[,.(Gene=gene,Variant=variant_qual,`Exonic part`=exonic_part)])
```

The gene lists used to select this section are available for review:

```{r gene_list}
knitr::kable(data.frame(Category=c("Activity-associated","Neuronal"),
  `Genes searched`=c(paste(activity_genes,collapse=", "),
    paste(neuronal_genes,collapse=", "))),check.names=FALSE)
```

Coordinates above follow the isoLASER output convention: variant positions
are zero-based (add 1 to compare with a VCF position); part intervals are
zero-based, with an exclusive end.

# Other top candidates and full results

All **`r nrow(gene_wide)` genes** are available in the interactive table below,
initially ranked by the largest **passing** absolute ΔPSI. Biological relevance
and read-level support should guide which candidates are investigated next.

```{r top_table}
top <- gene_wide[order(-max_signal)]
interactive_table(candidate_table(top))
```

All effect tables use percentage points. Download the
[complete gene summary](assets/roc_gene_signals_wide.tsv),
[selected events by timepoint](assets/roc_gene_tp_signals.tsv),
[all passing rows](assets/roc_cis_events_filtered.tsv.gz), and
[all associations across timepoints](assets/roc_event_wide.tsv.gz).

# Interpretation and next step

This screen provides candidates for allele-associated splicing. To determine
whether stimulation changes an allelic effect, compare reference- and
alternate-allele inclusion rates within the **same heterozygous donors** across
timepoints, show the supporting read counts, and inspect the top alignments.

The joint estimates have not yet been validated within donors, and no
multiple-testing correction has been applied. Detection can change with read
depth or donor composition. Variants absent from these four donors cannot be
tested. The [variant validation](isolaser_variants.html) measures concordance
among called sites; it does not estimate sensitivity.

The earlier GTEx ONT model analysis and model comparison are preliminary;
this page summarizes the default Nanopore model only. The current results are
generated by `9.isolaser_results/5.prepare_roc_results.R`.

Related pages: [variant validation](isolaser_variants.html),
[junction screen and sashimi plots](splicing_candidates.html),
[isoLASER setup](isolaser.html).
