---
title: "Genome-wide allele-specific splicing"
author: "XSun"
date: "2026-09-28"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      fig.width=7, fig.height=5)
library(data.table)
library(ggplot2)
library(ggrepel)

gene_tp <- fread("docs/assets/roc_gene_tp_signals.tsv")
gene_wide <- fread("docs/assets/roc_gene_signals_wide.tsv")
fmt <- function(n) format(n, big.mark=",", trim=TRUE)
sig <- function(x) !is.na(x) & x == TRUE
```

We ran [isoLASER](https://github.com/gxiaolab/isoLASER) jointly across four
donors on all 22 autosomes, at each of the three timepoints (0, 1, and 6 hours
after stimulation). isoLASER phases reads onto haplotypes using heterozygous
variants it discovers in the RNA, then tests whether the two copies of each gene
splice differently — allele-specific splicing, reported as **dPSI** (the
difference in Percent Spliced In between haplotypes). The dPSI reported here
is from the joint model across donors; per-donor allele-specific rates have not
been examined separately.

The model used is `--platform Nanopore`, which loads the default
`GM12878_ONT_ROC` classifier. A second ONT classifier (`GTEx_ONT_F1`) ships
with isoLASER but is currently commented out upstream; we tested both and found
F1 recovers more heterozygous sites, but use the default here pending
clarification from the developers.

# Filtering

Four filters were applied to the raw isoLASER cis calls:

1. **Effect size:** |dPSI| >= 0.10. Without this, many statistically significant
   calls have dPSI of 0.01–0.02, driven by high read depth rather than a
   meaningful splicing difference.

2. **Clustered variants excluded (conservative heuristic).** When multiple
   variants cluster within 30 bp, they may reflect sequencing errors in
   difficult sequence context. This is a conservative exclusion — some real
   variants in complex regions will also be removed. 34% of all variants fall
   in such clusters.

3. **Per-donor allelic balance.** At least one donor must have allelic balance
   0.25–0.75 and read depth >= 10 at the phasing variant. This is checked
   per donor, not averaged across donors — an average can hide a case where
   no individual donor is heterozygous (e.g., one reference-only and one
   alternate-only donor averaging to 0.5). This confirms phasing eligibility
   in at least one donor but does not verify within-donor allele-specific
   splicing.

4. **Minimum exonic part length >= 10 bp.** The exon model splits overlapping
   annotations into non-overlapping parts, which can create fragments of 1–3 bp.
   Splicing measured over such short regions is unreliable.

# Signal counts

After filtering: **`r fmt(uniqueN(gene_wide$gene))` genes** with at least one
event passing all filters. Each gene is represented by its strongest
(gene, variant, exonic part) combination, tracked as a fixed unit across all
three timepoints.

The bar chart below counts genes whose **selected representative event** passes
at each timepoint. Counts of genes with **any** passing event are higher (445,
576, and 626 at 0, 1, and 6 hours respectively).

```{r genes_per_tp, fig.height=4}
tp_summary <- gene_tp[sig(passes), .(n_genes=uniqueN(gene)), by=tp]
tp_summary[, tp := factor(tp, levels=c("0hr","1hr","6hr"))]

ggplot(tp_summary, aes(tp, n_genes)) +
  geom_col(fill=c("#888888","#E69F00","#D55E00"), width=0.6) +
  geom_text(aes(label=n_genes), vjust=-0.3, size=4) +
  labs(x="Timepoint", y="Genes (selected event passes)") +
  ylim(0, max(tp_summary$n_genes)*1.15) +
  theme_minimal(base_size=14)
```

# Timepoint patterns

For each gene's selected event, we classified its detection across the three
timepoints. The classification requires a **usable** baseline (at least one
heterozygous donor with sufficient depth at 0 hr for that event) before
comparing timepoints.

```{r tp_patterns, fig.height=4}
pat <- gene_wide[, .N, by=pattern]
pat[, pattern := factor(pattern, levels=c("detected_after_stim",
                                           "post_stim_only",
                                           "all_timepoints",
                                           "detected_at_baseline_only","other"))]
setorder(pat, pattern)

ggplot(pat, aes(pattern, N)) +
  geom_col(fill="#4477AA", width=0.6) +
  geom_text(aes(label=N), vjust=-0.3, size=3.5) +
  labs(x="", y="Number of genes") +
  theme_minimal(base_size=13) +
  theme(axis.text.x=element_text(angle=30, hjust=1))
```

```{r tp_table}
pat_desc <- data.frame(
  Pattern=c("detected_after_stim", "post_stim_only",
            "all_timepoints", "detected_at_baseline_only", "other"),
  Description=c(
    "Usable at 0 hr (het donor present) with no passing signal; passes after stimulation",
    "No usable baseline (no het donor at 0 hr); signal only at 1 or 6 hr — absence at baseline is not established",
    "Passes at all three timepoints (constitutive)",
    "Passes at 0 hr but not after stimulation, with usable data at the later timepoints",
    "Does not fit the above categories"),
  Genes=c(sum(gene_wide$pattern=="detected_after_stim"),
          sum(gene_wide$pattern=="post_stim_only"),
          sum(gene_wide$pattern=="all_timepoints"),
          sum(gene_wide$pattern=="detected_at_baseline_only"),
          sum(gene_wide$pattern=="other")))
knitr::kable(pat_desc)
```

**`r sum(gene_wide$pattern=="detected_after_stim")` genes** had a usable
baseline with no passing signal, then passed filters after stimulation. These
are differences in signal detection between timepoints; establishing that the
underlying allele-specific effect itself changes with stimulation would require
comparing allele-specific splicing rates within individual heterozygous donors
at each timepoint.

The remaining `r sum(gene_wide$pattern=="post_stim_only")` genes had no usable
baseline data for the selected event (no heterozygous donor at 0 hr), so
baseline absence cannot be assessed.

# Top signals

The table shows measured |dPSI| at each timepoint. NA means the event was not
present in the data; values below the 0.10 threshold or failing other filters
are shown as measured (not zeroed out). The "passes" column indicates which
timepoints passed all filters.

```{r top_table}
top_tab <- gene_wide[order(-max_signal)][1:30]
top_tab[, passes := paste0(
  fifelse(sig(passes_0hr), "0", ""),
  fifelse(sig(passes_1hr), "1", ""),
  fifelse(sig(passes_6hr), "6", ""))]
knitr::kable(top_tab[, .(gene,
                          dpsi_0hr=round(abs(dpsi_0hr),3),
                          dpsi_1hr=round(abs(dpsi_1hr),3),
                          dpsi_6hr=round(abs(dpsi_6hr),3),
                          passes, pattern)],
             col.names=c("Gene","|dPSI| 0 hr","|dPSI| 1 hr","|dPSI| 6 hr",
                          "Passes (hr)","Pattern"))
```

```{r top30, fig.height=9, fig.width=7}
top_genes <- gene_wide[order(-max_signal)][1:30, gene]

top_data <- gene_tp[gene %in% top_genes & sig(passes)]
top_data[, tp := factor(tp, levels=c("0hr","1hr","6hr"))]
top_data[, gene := factor(gene, levels=rev(top_genes))]

ggplot(top_data, aes(tp, gene, size=abs_dpsi, color=abs_dpsi)) +
  geom_point() +
  scale_size_continuous(range=c(1,8), name="|dPSI|") +
  scale_color_gradient(low="#DDDDDD", high="#CC0000", name="|dPSI|") +
  labs(x="Timepoint", y="") +
  theme_minimal(base_size=12) +
  theme(panel.grid.major=element_line(color="#EEEEEE"))
```

# Neuronal and activity-regulated genes

```{r named_genes, fig.height=6, fig.width=8}
ieg <- c("FOS","FOSB","FOSL1","FOSL2","JUN","JUNB","JUND",
         "EGR1","EGR2","EGR3","EGR4","NR4A1","NR4A2","NR4A3",
         "ARC","BDNF","NPAS4","SIK1","ATF3","GADD45B","GADD45G",
         "HOMER1","IER2","IER5","BTG2","DUSP1","DUSP5","DUSP6",
         "NFKBIA","NFKBIZ","CREM","CYR61","CTGF","TIPARP","PPP1R15A",
         "ZFP36","ZFP36L1","BHLHE40","BHLHE41","KLF2","KLF4",
         "RGS2","SERTAD1","CITED2")
neuronal <- c("NRXN1","NRXN2","NRXN3","NLGN1","NLGN2","NLGN3",
              "GRIN1","GRIN2A","GRIN2B","GRIA1","GRIA2","GRIA3","GRIA4",
              "DLG4","SHANK1","SHANK2","SHANK3","CAMK2A","CAMK2B",
              "SYN1","SYT1","SNAP25","CACNA1A","CACNA1B","CACNA1C","CACNA1E",
              "SCN1A","SCN2A","SCN8A","CELF4","RBFOX1","RBFOX2","RBFOX3",
              "NOVA1","NOVA2","PTBP2","ANK3","MAPT","APP","DSCAM",
              "CNTNAP2","CADM1","TTLL7","CLDND1","MEF2C")

gene_all <- gene_tp[sig(passes), .(max_dpsi=max(abs_dpsi), mean_cov=median(COV, na.rm=TRUE)), by=gene]
gene_all <- merge(gene_all, gene_wide[, .(gene, pattern)], by="gene")
gene_all[, label_cat := fcase(
  gene %in% ieg, "Activity-regulated",
  gene %in% neuronal, "Neuronal",
  default = "Other"
)]
gene_all[, show_label := label_cat != "Other" & max_dpsi >= 0.10]

ggplot(gene_all, aes(mean_cov, max_dpsi, color=label_cat)) +
  geom_point(data=gene_all[label_cat=="Other"], alpha=0.3, size=1.5) +
  geom_point(data=gene_all[label_cat!="Other"], alpha=0.8, size=3) +
  geom_text_repel(data=gene_all[show_label==TRUE],
                  aes(label=gene), size=3.5, max.overlaps=20, show.legend=FALSE) +
  scale_color_manual(values=c("Activity-regulated"="#D55E00",
                               "Neuronal"="#0072B2", "Other"="#BBBBBB"), name="") +
  scale_x_log10() +
  labs(x="Median coverage (reads)", y="Max |dPSI| across timepoints") +
  theme_minimal(base_size=13)
```

Labeled genes are from curated lists of activity-regulated immediate-early genes
(orange) and neuronal-function genes (blue).

```{r named_table}
named <- gene_wide[gene %in% c(ieg, neuronal)]
named[, category := fifelse(gene %in% ieg, "Activity-regulated", "Neuronal")]
named <- named[max_signal > 0]
setorder(named, -max_signal)
named[, passes := paste0(
  fifelse(sig(passes_0hr), "0", ""),
  fifelse(sig(passes_1hr), "1", ""),
  fifelse(sig(passes_6hr), "6", ""))]
knitr::kable(named[, .(gene, category,
                        dpsi_0hr=round(abs(dpsi_0hr),3),
                        dpsi_1hr=round(abs(dpsi_1hr),3),
                        dpsi_6hr=round(abs(dpsi_6hr),3),
                        passes, pattern)],
             col.names=c("Gene","Category","|dPSI| 0 hr","|dPSI| 1 hr",
                          "|dPSI| 6 hr","Passes (hr)","Pattern"))
```

Many classic immediate-early genes (JUN, JUNB, JUND, IER2) are single-exon
genes with no splicing to test. Others (e.g., FOS) have isoLASER output
rows but do not pass the effect-size or quality filters at any timepoint.

# Top neuronal gene plots

```{r gene_plots, fig.height=4, fig.width=7, results='asis'}
event_data <- fread("/project/xinhe/xsun/long_read_RNA/9.isolaser_results/out/roc_event_wide.tsv.gz")
focus_genes <- c("GRIA2","ANK3","CACNA1E","SCN2A","CAMK2B","SYT1","NFKBIA")
focus_genes <- focus_genes[focus_genes %in% event_data$gene]

for(g in focus_genes) {
  gd <- event_data[gene == g & pattern != "other"]
  if(nrow(gd) == 0) next
  gd_long <- melt(gd, id.vars=c("gene","variant_qual","exonic_part","pattern"),
                  measure.vars=c("dpsi_0hr","dpsi_1hr","dpsi_6hr"),
                  variable.name="tp", value.name="dpsi")
  gd_long[, tp := factor(sub("dpsi_","",tp), levels=c("0hr","1hr","6hr"))]
  gd_long <- gd_long[!is.na(dpsi)]

  gd_long[, event_id := paste0(
    sub(".*:(\\d+):.*","\\1", variant_qual), " / ",
    sub("chr[0-9]+:(\\d+)-(\\d+):.*","\\1-\\2", exonic_part))]

  top_events <- gd[order(-max_passing_dpsi)][1:min(.N, 6), .(variant_qual, exonic_part)]
  gd_long <- gd_long[top_events, on=.(variant_qual, exonic_part)]

  p <- ggplot(gd_long, aes(tp, dpsi)) +
    geom_hline(yintercept=0, lty=2, color="grey60") +
    geom_point(size=3, color="#0072B2") +
    geom_line(aes(group=1), color="#0072B2", alpha=0.5) +
    facet_wrap(~event_id, scales="free_y") +
    labs(x="Timepoint", y="dPSI", title=g) +
    theme_minimal(base_size=12)
  print(p)
  cat("\n\n")
}
```

Each panel shows one (variant, exonic part) combination tracked across
timepoints. The variant position and exonic part coordinates are in the facet
label (variant position / part start-end). These are joint dPSI values across
donors; per-donor allele-specific rates would be needed to confirm the signal
in individual heterozygous donors.

# Interpretation

This analysis identifies **`r nrow(gene_wide)` genes** with at least one event
passing quality filters. Each gene is represented by a fixed (variant, exonic
part) combination tracked across all three timepoints.

**`r sum(gene_wide$pattern=="detected_after_stim")` genes** had usable data at
baseline (a heterozygous donor present) with no passing signal, and then passed
filters after stimulation. These are differences in signal detection between
timepoints. Whether the underlying allele-specific splicing effect itself
changes with stimulation — as opposed to changes in power, expression level,
or isoform composition — would require comparing reference-versus-alternate
splicing rates within individual heterozygous donors at each timepoint.

Another `r sum(gene_wide$pattern=="post_stim_only")` genes had no usable
baseline data for the selected event, so baseline absence cannot be assessed.

Caveats:

- **Joint model only.** The dPSI values are from isoLASER's joint analysis
  across donors. Within-donor allele-specific splicing rates have not been
  examined.
- **Concordance, not sensitivity.** The
  [variant validation](isolaser_variants.html) page measures agreement between
  isoLASER calls and the genotype panel; it does not estimate what fraction of
  true variants isoLASER detects, because it starts from isoLASER's calls.
- **Four donors.** With n = 4, a signal requires a heterozygous variant in at
  least one donor. Variants absent from these donors cannot be tested.
- **No multiple-testing correction** has been applied. The p-values from
  isoLASER's joint model are shown as-is.

Related pages: [variant validation](isolaser_variants.html),
[junction screen](splicing_candidates.html),
[isoLASER setup](isolaser.html).
