---
title: "Genome-wide allele-specific splicing"
author: "XSun"
date: "2026-09-28"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      fig.width=7, fig.height=4)
library(data.table)
library(ggplot2)
library(DT)
gene_tp <- fread("docs/assets/roc_gene_tp_signals.tsv")
gene_wide <- fread("docs/assets/roc_gene_signals_wide.tsv")
events <- fread("docs/assets/roc_event_wide.tsv.gz")
passing <- fread("docs/assets/roc_cis_events_filtered.tsv.gz")
joint_vs_het <- fread("docs/assets/roc_joint_vs_het.tsv")
joint_vs_het_lost <- fread("docs/assets/roc_joint_vs_het_lost.tsv")
depth_summary <- fread("docs/assets/depth_summary.tsv")
depth_reasons <- fread("docs/assets/depth_change_reasons.tsv")
depth_effect <- fread("docs/assets/depth_effect.tsv")
depth_genes <- fread("docs/assets/depth_gene_status.tsv")
sig <- function(x) !is.na(x) & x == TRUE
fmt <- function(n) format(n, big.mark=",", trim=TRUE)
tp_levels <- c("0hr", "1hr", "6hr")
pattern_names <- c(detected_after_stim="Detected after stimulation",
  post_stim_only="Baseline unavailable", all_timepoints="All three timepoints",
  detected_at_baseline_only="Baseline only", other="Other detection patterns")
pass_times <- function(d) vapply(seq_len(nrow(d)), function(i) {
  keep <- sig(unlist(d[i, .(passes_0hr, passes_1hr, passes_6hr)]))
  paste(c("0", "1", "6")[keep], collapse=", ")
}, character(1))
effect_table <- function(d) {
  data.table(Gene=d$gene,
    `0 hr`=round(100*abs(d$dpsi_0hr),1),
    `1 hr`=round(100*abs(d$dpsi_1hr),1),
    `6 hr`=round(100*abs(d$dpsi_6hr),1),
    `Passing timepoints (hr)`=pass_times(d),
    Pattern=unname(pattern_names[d$pattern]))
}
interactive_table <- function(d) DT::datatable(as.data.frame(d), rownames=FALSE,
  filter="top", extensions="Buttons", options=list(
    pageLength=10, lengthMenu=c(10,25,50,100), scrollX=TRUE,
    order=list(), dom="Blfrtip", buttons=c("copy","csv")),
  caption="Search, filter, or click a column heading to sort; export the filtered table as CSV.")
neuronal_genes <- c("NRXN1","NRXN2","NRXN3","NLGN1","NLGN2","NLGN3",
  "GRIN1","GRIN2A","GRIN2B","GRIA1","GRIA2","GRIA3","GRIA4",
  "DLG4","SHANK1","SHANK2","SHANK3","CAMK2A","CAMK2B","SYN1","SYT1","SNAP25",
  "CACNA1A","CACNA1B","CACNA1C","CACNA1E","SCN1A","SCN2A","SCN8A",
  "CELF4","RBFOX1","RBFOX2","RBFOX3","NOVA1","NOVA2","PTBP2","ANK3","MAPT",
  "APP","DSCAM","CNTNAP2","CADM1","TTLL7","MEF2C","DCLK1","DTNBP1","CADPS",
  "DOCK7","DMXL2","SIGMAR1","TMEM106B","FNDC5","ELAVL4","PARK7")
activity_genes <- c("FOS","FOSB","FOSL1","FOSL2","JUN","JUNB","JUND",
  "EGR1","EGR2","EGR3","EGR4","NR4A1","NR4A2","NR4A3","ARC","BDNF","NPAS4",
  "SIK1","ATF3","GADD45B","GADD45G","HOMER1","IER2","IER5","BTG2","DUSP1",
  "DUSP5","DUSP6","NFKBIA","NFKBIZ","CREM","TIPARP","PPP1R15A","ZFP36",
  "BHLHE40","BHLHE41","KLF2","KLF4","RGS2","SERTAD1","CITED2")
neuronal_events <- copy(events[gene %in% c(neuronal_genes,activity_genes)])
priority_patterns <- c("detected_after_stim","all_timepoints",
  "detected_at_baseline_only","other","post_stim_only")
neuronal_events[, priority := match(pattern,priority_patterns)]
setorder(neuronal_events,priority,-max_passing_dpsi)
neuronal_selected <- neuronal_events[, .SD[1], by=gene]
setorder(neuronal_selected,priority,-max_passing_dpsi)
candidate_table <- function(d) {
  ans <- effect_table(d)
  ans[, Variant := d$variant_qual]
  ans[, `Exonic part` := d$exonic_part]
  ans
}
```

We screened **22 autosomes, four donors, and three timepoints (0, 1, and
6 hours)** using isoLASER's default Nanopore model (`GM12878_ONT_ROC`).
isoLASER's joint step also pools homozygous donors, so we recalculated every
association from **heterozygous donors only**. This identifies
**`r fmt(nrow(gene_wide))` genes** with candidate allele-associated splicing
signals. These are candidates for follow-up; agreement between individual
donors and stimulation dependence have not yet been confirmed.

# How the screen works

For each donor and timepoint, isoLASER counts the reads that cover both a
variant and an **exonic part** (a segment of the gene's exon structure). It
sorts them into a 2×2 table: reference or alternate allele, by including or
excluding the part.

isoLASER's joint step rescales each donor's table to equal weight, adds the
tables together, and tests the pooled table. It includes every donor with at
least 10 reads, **including homozygous donors**. A homozygous-reference donor
adds only reference-allele reads, so the pooled difference can reflect
differences between donors rather than between the two alleles.

We rebuilt each pooled table from **heterozygous donors only**. The rescaling,
association score and `cis` threshold are isoLASER's own functions, so donor
selection is the only change. As a check, the same code with all donors
reproduces isoLASER's output exactly in 167,773 of the 167,782 rows we could
check; the 9 exceptions are one KPNB1 variant, and none pass the filters.

We retain associations that meet all of these criteria:

| Filter | Requirement | Purpose |
|:--|:--|:--|
| Heterozygous donors | Alternate-allele fraction 0.25–0.75 and ≥30 reads; only these donors are pooled, and at least one is required | Compare the two alleles within the same people |
| Association | isoLASER's `cis` label, recalculated on the heterozygous-donor table (each allele needs ≥10 reads) | Association above isoLASER's null threshold |
| Allelic difference | \(\lvert\Delta\mathrm{PSI}\rvert \geq 0.10\) | At least a 10 percentage-point difference |
| Part length | ≥10 bases | Exclude very short fragments |
| Variant spacing | No other tested variant within 30 bases | Conservative exclusion of clustered calls; real variants can also be removed |

For timepoint comparisons, we keep the **same gene, variant, and exonic part**.
The representative event for each gene is the one with the largest effect
that passes these filters. A usable timepoint has an output row and at least
one heterozygous donor with ≥30 reads; the heterozygous donors need not be the
same at different timepoints.

isoLASER itself requires 10 reads per pooled donor; we use 30 as a
conservative choice. In the [variant validation](isolaser_variants.html),
isoLASER labels panel heterozygotes as homozygous less often at 30 or more
reads (21% of calls) than at 20 or more (29%). These calculations count each
read by the allele it carries and do not use those labels, and allele balance
alone identified panel heterozygotes correctly in 90–95% of calls at every
depth from 10 reads (calculated from the validation call table). The 30-read
cutoff therefore mainly removes shallow donors, whose estimates are noisier,
and keeps about 85% of the associations and genes found at 20 reads. [Effect of the depth cutoff](#effect-of-the-depth-cutoff) compares
the results at other cutoffs.

## Effect of using heterozygous donors only

Applying the same filters to isoLASER's original pooled values and to the
heterozygous-donor values:

```{r joint_vs_het}
knitr::kable(data.frame(Method=joint_vs_het$method, `Passing rows`=fmt(joint_vs_het$rows),
  Associations=fmt(joint_vs_het$associations), Genes=fmt(joint_vs_het$genes),
  check.names=FALSE))
```

Of the rows that pass with isoLASER's pooling but not with heterozygous donors
only, `r fmt(joint_vs_het_lost[reason=="no longer cis", N])` are no longer
`cis`, `r fmt(joint_vs_het_lost[reason=="too few het reads (<10 per allele)", N])`
have fewer than 10 heterozygous-donor reads on one allele, and
`r fmt(joint_vs_het_lost[reason=="|dPSI| below 0.10", N])` fall below the
effect-size cutoff. A smaller pooled table has less power, so some of these
losses are real associations that can no longer be detected, not artifacts.
The change works in both directions:

- **SCN2A, 1 hr:** one heterozygous donor and three homozygous-reference
  donors. The pooled difference is 12.0 points and `cis`; the heterozygous
  donor alone gives 6.5 points and is not `cis`.
- **RPL8, 0 hr:** one heterozygous donor and three homozygous donors, which
  dilute the effect. The pooled difference is 0.3 points and not `cis`; the
  heterozygous donor alone gives 47.9 points and is `cis`.
- **RHOC, 0 hr:** all four donors are heterozygous, and both calculations give
  78.8 points.

## What the numbers mean

For a given part, an allele-specific inclusion rate is:

\[
\mathrm{PSI}_{\mathrm{REF}}=
\frac{\text{reference-allele reads including the part}}
{\text{reference-allele reads including or excluding the part}},
\qquad
\Delta\mathrm{PSI}=\mathrm{PSI}_{\mathrm{REF}}-\mathrm{PSI}_{\mathrm{ALT}}.
\]

For example, 16/20 reference reads versus 6/20 alternate reads gives
80% versus 30%, or an inclusion difference of **50 percentage points**.
This is an illustration, not a measured result from this dataset.

**Sign convention in these outputs:** the installed joint implementation
computes its signed `delta_psi(ref-alt)` from the first count category, which
is the part's **exclusion** category. Thus the reported joint difference is
reference exclusion rate minus alternate exclusion rate; the inclusion
difference has the opposite sign. The absolute effect is unchanged. The
pooled calculation rescales each heterozygous donor to equal weight rather
than summing raw reads.

# How many signals were found?

```{r count_table}
knitr::kable(data.frame(
  Unit=c("Passing rows", "Distinct gene–variant–part combinations", "Genes"),
  Count=c(fmt(nrow(passing)), fmt(nrow(events)), fmt(nrow(gene_wide))),
  Meaning=c("One gene–variant–exonic-part association at one timepoint",
    "Associations combined across timepoints", "At least one passing association")))
```

These units are not independent biological events: multiple variants or
adjacent parts may describe the same underlying splicing change.
For example, one gene–variant–part association passing at 0, 1, and 6 hours
contributes **three passing rows**, but only **one distinct association**.

```{r genes_per_tp}
tp_summary <- passing[, .(Genes=uniqueN(gene)), by=tp]
tp_summary[, tp := factor(tp, levels=tp_levels)]
ggplot(tp_summary, aes(tp, Genes, fill=tp)) +
  geom_col(width=0.6, show.legend=FALSE) +
  geom_text(aes(label=Genes), vjust=-0.4, size=4) +
  scale_fill_manual(values=c("0hr"="#888888", "1hr"="#E69F00", "6hr"="#D55E00")) +
  scale_x_discrete(labels=c("0 hr", "1 hr", "6 hr")) +
  scale_y_continuous(expand=expansion(mult=c(0,0.12))) +
  labs(x=NULL, y="Genes with any passing association") +
  theme_minimal(base_size=13)
```

## Effect of the depth cutoff

The depth cutoff applies to each donor separately: a heterozygous donor is
pooled only if it has at least that many reads covering both the variant and
the exonic part. We reran the heterozygous-donor screen at 10 reads (isoLASER's
own minimum), 20, 30 (used throughout this page) and 50.

```{r depth_table}
depth_order <- c(dp10="10 reads", dp20="20 reads", dp30="30 reads (this page)",
  dp50="50 reads")
ds <- depth_summary[match(names(depth_order), screen)]
vs_base <- function(x) ifelse(ds$screen == "dp30", "–", fmt(x))
knitr::kable(data.frame(Cutoff=unname(depth_order), `Passing rows`=fmt(ds$rows),
  Associations=fmt(ds$associations), Genes=fmt(ds$genes),
  `Genes kept`=vs_base(ds$genes_kept), `Genes lost`=vs_base(ds$genes_lost),
  `Genes gained`=vs_base(ds$genes_gained), check.names=FALSE), align="lrrrrrr")
```

**Column guide**

- **Passing rows:** gene–variant–exonic-part associations that pass all
  filters at one timepoint. An association passing at 0, 1 and 6 hours counts
  three times.
- **Associations:** distinct gene–variant–exonic-part combinations that pass
  at one or more timepoints, so each counts once.
- **Genes:** genes with at least one passing association.
- **Genes kept / lost / gained:** compared with the 30-read screen. Kept genes
  pass in both screens, lost genes pass only at 30 reads, and gained genes pass
  only at the other cutoff.

```{r depth_reason_values}
reason_pct <- function(s, ch, r) {
  v <- depth_reasons[screen == s & change == ch & reason == r, pct]
  if (length(v)) v else 0
}
```

- **Lowering the cutoff to 10 or 20 reads** adds associations supported by
  shallower donors. Of the rows gained,
  `r reason_pct("dp10", "gained", "no het donor meets the cutoff")`% (10 reads)
  and `r reason_pct("dp20", "gained", "no het donor meets the cutoff")`% (20 reads)
  have no heterozygous donor with 30 reads; most of the rest are not `cis`
  once the shallow donors are removed. A few associations are lost, because
  adding shallow donors to the pool dilutes them.
- **Raising the cutoff to 50 reads** mostly removes associations in which no
  heterozygous donor reaches 50 reads
  (`r reason_pct("dp50", "lost", "no het donor meets the cutoff")`% of the rows
  lost); most of the rest are no longer `cis`. The few gained rows were not
  `cis` at 30 reads and pass once a shallower donor is removed.

The next table asks whether the associations that change with the cutoff
differ in effect size. For each cutoff, associations are split into three
groups compared with the 30-read screen: **kept** (pass in both screens),
**lost** (pass only at 30 reads) and **gained** (pass only at the other
cutoff). Each cell shows the median, across the group, of each association's
largest passing |ΔPSI| in percentage points, with the number of associations
in parentheses. Kept and lost effects are measured at 30 reads; gained effects
at the other cutoff.

```{r depth_effect_table}
de <- dcast(depth_effect, screen ~ status, value.var=c("median_abs_dpsi", "n"))
de <- de[match(setdiff(names(depth_order), "dp30"), screen)]
cell <- function(m, n) paste0(round(100 * m, 1), " (", fmt(n), ")")
knitr::kable(data.frame(Cutoff=unname(depth_order[de$screen]),
  Kept=cell(de$median_abs_dpsi_kept, de$n_kept),
  Lost=cell(de$median_abs_dpsi_lost, de$n_lost),
  Gained=cell(de$median_abs_dpsi_gained, de$n_gained), check.names=FALSE),
  align="lrrr")
```

The associations that depend on shallower donors (gained at 10 and 20 reads,
lost at 50 reads) have larger effects than those that pass regardless.
With fewer reads, only larger differences reach the `cis` threshold, and small
samples also tend to overestimate effects, so associations that rely on
shallow donors are less certain.

**`r fmt(depth_genes[, sum(dp10 & dp20 & dp30 & dp50)])` genes** pass at every
cutoff; these are the candidates least sensitive to the choice. The
[gene status table](assets/depth_gene_status.tsv) marks the screens in which
each gene passes. Scripts: `9.isolaser_results/6.het_only_stats.py` (run once
per cutoff), `5.prepare_roc_results.R` (for this page,
`Rscript 5.prepare_roc_results.R 30 out/dp30 0`) and
`7.compare_depth_thresholds.R`.

## Detection across timepoints

The following counts describe each gene's **selected representative event**.
They describe filter outcomes, not a statistical test of whether stimulation
changes the allelic effect.

**Usable** means that this exact gene–variant–part combination has a joint
output row at that timepoint, and at least one heterozygous donor
(alternate-allele fraction **0.25–0.75**, **≥30 reads**) for the combination.
Usable data can still fail: the heterozygous-donor table may have fewer than
10 reads on one allele, or miss the `cis` or effect-size criteria. **No usable
baseline** can mean a missing output row, insufficient depth, or no
heterozygous donor; it does not establish that the gene is unexpressed.

The pattern **Baseline unavailable** is exactly this case: the association
passes at 1 or 6 hours, but 0 hours has no usable data, so we cannot tell
whether the association was already present before stimulation. Every other
pattern has a usable baseline.

```{r detection_barplot, fig.height=4.5}
detection_counts <- data.table(pattern=names(pattern_names))
detection_counts[, Genes := vapply(pattern,function(p)sum(gene_wide$pattern==p),integer(1))]
detection_counts[, Pattern := factor(unname(pattern_names[pattern]),
  levels=rev(unname(pattern_names)))]
ggplot(detection_counts,aes(Pattern,Genes,fill=pattern)) +
  geom_col(width=0.65,show.legend=FALSE) +
  geom_text(aes(label=Genes),hjust=-0.2,size=4) + coord_flip() +
  scale_fill_manual(values=c(detected_after_stim="#0072B2",post_stim_only="#BBBBBB",
    all_timepoints="#009E73",detected_at_baseline_only="#E69F00",other="#999999")) +
  scale_y_continuous(expand=expansion(mult=c(0,0.12))) +
  labs(x=NULL,y="Genes (selected representative event)") + theme_minimal(base_size=12)
```

```{r detection_table}
patterns <- names(pattern_names)
descriptions <- c(
  "Usable baseline does not pass; passes at 1 or 6 hr",
  "No usable baseline; passes at 1 or 6 hr",
  "Passes at 0, 1, and 6 hr",
  "Passes at 0 hr; usable at 1 and 6 hr but passes at neither",
  "Mixed detection or incomplete usable timepoints")
knitr::kable(data.frame(Pattern=unname(pattern_names),
  Genes=vapply(patterns,function(p)sum(gene_wide$pattern==p),integer(1)),
  Definition=descriptions), row.names=FALSE)
```

**`r sum(gene_wide$pattern=="detected_after_stim")` genes** pass after
stimulation despite a usable, nonpassing baseline. They are a useful starting
set for follow-up. For **`r sum(gene_wide$pattern=="post_stim_only")` genes**,
baseline absence cannot be assessed.

# Neuronal and activity-associated candidates

The table below expands the examples to **`r nrow(neuronal_selected)` genes**
from a manually selected list of neuronal and activity-associated genes.
**`r sum(neuronal_selected$pattern=="detected_after_stim")` have an association
detected after stimulation with a usable, nonpassing baseline.** These appear
first, followed by other baseline-usable patterns; candidates without a usable
baseline appear last. Biological relevance helps prioritize examples but does
not establish a stimulation-dependent allelic effect.

For this section, we search **all passing associations**, rather than only the
strongest event per gene. We select one association per gene by detection
category, then by the largest passing effect within that category. This can
select a different event from the genome-wide representative above. Each
selected variant–part combination remains fixed across timepoints; the
genome-wide counts and detection categories above are unchanged.

The candidates include [PARK7](https://www.ncbi.nlm.nih.gov/gene/11315), a
Parkinson's disease gene; [MAPT](https://www.ncbi.nlm.nih.gov/gene/4137),
which encodes tau; [APP](https://www.ncbi.nlm.nih.gov/gene/351);
[SYT1](https://www.ncbi.nlm.nih.gov/gene/6857), a calcium sensor for
neurotransmitter release; and the activity-induced gene
[NR4A1](https://www.ncbi.nlm.nih.gov/gene/3164). Several of these rest on a
single heterozygous donor, so they are not yet replicated across donors. The
**Pattern** column uses the detection categories defined above.

```{r examples_table}
neuronal_table <- candidate_table(neuronal_selected)
neuronal_table[, Category := ifelse(neuronal_selected$gene %in% activity_genes,
  "Activity-associated","Neuronal")]
interactive_table(neuronal_table)
```

Table values are **absolute allelic differences in percentage points**, not
the percentage of transcripts including an exon. `NA` means no
heterozygous-donor estimate was available. Nonpassing measured effects remain visible; the passing
timepoints are listed separately.

## Plots for all neuronal candidates

The plots show **all `r nrow(neuronal_selected)` genes in the table above**,
using exactly the same selected variant–part combination and the same gene
order. Candidates detected after stimulation with a usable baseline appear
first; other detection patterns, including unavailable baselines, follow.
No estimate is substituted for missing data. Read-level inspection and
within-donor comparisons are the next steps for these examples.

```{r example_trends, fig.height=14, fig.width=9}
focus <- copy(neuronal_selected)
plot_data <- melt(focus,id.vars=c("gene","variant_qual","exonic_part"),
  measure.vars=list(dpsi=paste0("dpsi_",tp_levels),
    passes=paste0("passes_",tp_levels),usable=paste0("usable_",tp_levels)),
  variable.name="time_index")
plot_data[, tp := tp_levels[as.integer(time_index)]]
plot_data[, tp := factor(tp,levels=tp_levels)]
plot_data[, gene := factor(gene,levels=focus$gene)]
plot_data[, Status := fifelse(sig(passes), "Passes filters",
  fifelse(sig(usable), "Usable; does not pass", "Does not meet donor eligibility"))]
plot_data <- plot_data[!is.na(dpsi)]
ggplot(plot_data,aes(tp,100*dpsi)) +
  geom_hline(yintercept=0,linetype=2,color="grey65") +
  geom_point(aes(shape=Status,color=Status),size=3) +
  facet_wrap(~gene,ncol=3) +
  scale_shape_manual(values=c("Passes filters"=16,"Usable; does not pass"=1,
    "Does not meet donor eligibility"=4)) +
  scale_color_manual(values=c("Passes filters"="#0072B2",
    "Usable; does not pass"="#777777","Does not meet donor eligibility"="#BBBBBB")) +
  scale_x_discrete(labels=c("0 hr","1 hr","6 hr"),drop=FALSE) +
  labs(x=NULL,y="Heterozygous-donor exclusion difference\n(REF − ALT, percentage points)",
    shape=NULL,color=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
```

Each panel follows the same association across timepoints on a common vertical
scale; a positive value means more exclusion on the reference allele.
Missing points are missing estimates. Each point pools the heterozygous donors
at that timepoint; individual donors' rates are not shown.

```{r example_coordinates}
knitr::kable(focus[,.(Gene=gene,Variant=variant_qual,`Exonic part`=exonic_part)])
```

The gene lists used to select this section are available for review:

```{r gene_list}
knitr::kable(data.frame(Category=c("Activity-associated","Neuronal"),
  `Genes searched`=c(paste(activity_genes,collapse=", "),
    paste(neuronal_genes,collapse=", "))),check.names=FALSE)
```

Coordinates above follow the isoLASER output convention: variant positions
are zero-based (add 1 to compare with a VCF position); part intervals are
zero-based, with an exclusive end.

# Other top candidates and full results

All **`r nrow(gene_wide)` genes** are available in the interactive table below,
initially ranked by the largest **passing** absolute ΔPSI. Biological relevance
and read-level support should guide which candidates are investigated next.

```{r top_table}
top <- gene_wide[order(-max_signal)]
interactive_table(candidate_table(top))
```

All effect tables use percentage points. Download the
[complete gene summary](assets/roc_gene_signals_wide.tsv),
[selected events by timepoint](assets/roc_gene_tp_signals.tsv),
[all passing rows](assets/roc_cis_events_filtered.tsv.gz), and
[all associations across timepoints](assets/roc_event_wide.tsv.gz).

# Interpretation and next step

This screen provides candidates for allele-associated splicing. To determine
whether stimulation changes an allelic effect, compare reference- and
alternate-allele inclusion rates within the **same heterozygous donors** across
timepoints, show the supporting read counts, and inspect the top alignments.

The estimates pool heterozygous donors; whether individual donors agree has
not been checked, and no multiple-testing correction has been applied. Detection can change with read
depth or donor composition. Variants absent from these four donors cannot be
tested. The [variant validation](isolaser_variants.html) measures concordance
among called sites; it does not estimate sensitivity.

The earlier GTEx ONT model analysis and model comparison are preliminary;
this page summarizes the default Nanopore model only. The heterozygous-donor
statistics are calculated by `9.isolaser_results/6.het_only_stats.py` (run via
`6.het_only_stats.sbatch`), and the filtered results by
`9.isolaser_results/5.prepare_roc_results.R`.

Related pages: [variant validation](isolaser_variants.html),
[junction screen and sashimi plots](splicing_candidates.html),
[isoLASER setup](isolaser.html).
